Trygve's Problem

### 2018-03-29 generativemodelsforscrna/outline.tex
new residual plots

\subsection{Posterior predictive means}

The posterior predictive mean can be interpreted as a ``denoised'' version of the original.  In this section we examine the Allen data from that point view.

For example, here we have selected 1000 random entries from the gene by cell matrix.  For each entry, we plot a point in which the horizontal location indicates the true value of that entry, and the vertical position indicates the posterior predictive expectation of that entry.

\includegraphics[width=0.8\textwidth]{pics/resids2}

We can compare this with a similar plot using posterior predictive samples instead of the true values.  This then forms a different kind of posterior predictive check.

\includegraphics[width=0.8\textwidth]{pics/resids3}

The fact that these two plots look the same is encouraging.

We can also use heatmaps to try to give more structure to our understanding of this ``denoising'' process.  Here we have three plots: one giving the original gene expression for the first 1000 cells, one giving the posterior expected gene expression for the same cells, and one giving the residual.  In all cases we have transformed the data by taking $\log(1+x)$ and subtracting gene-level means.  Each column corresponds to a gene, and each row corresponds to a cell.  We have sorted the cells by one of the latent components, and sorted the genes by using a property of the gene decoder.

@@

Note that there is a group of cells which are much more difficult to understand, which are found at the bottom of the plot.  These are mostly astrocytes.

Since we have 20000 cells, it might be nice to make a similar plot for all the cells.  However, such a plot would be very tall and perhaps difficult to look at.  We can instead take the full 20000 x 1400 plot and average neighboring rows and columns together.  This gives a plot like this:

\includegraphics[width=0.6\textwidth]{pics/residsfull}

If there was a lot of structure in the residual, we would feel that perhaps there was something important and obvious we were missing.  Since there isn't, we are reasonably encouraged.

>>>>

Another way to see that the model isn't quite right is to look at posterior predictive p values.  That is, for each entry of the matrix, we can compute the cumulative distribution function of that entry in light of the parameters predicted by the posterior mode.  Here we have done that for the first 1000 entries; the results are not exactly encouraging:

@@

We see that for some genes the $p$ values (which should be distributed uniformly) are almost always near 1, and in other cases they are almost always near 0.  We can dig deeper into this by looking at individual genes, and comparing the expected value for that gene under the model with the actual value.  There are clearly some systematic differences.

\includegraphics[width=0.8\textwidth]{pics/residsdetail}

One potential reason for this might be that the negative binomial data distribution is inappropriate.  To diagnose this, we fitted a zero-inflated negative binomial distribution to each collection of cells Allen Institute dubbed a ``cluster.''  We then measured the goodness-of-fit, in terms of whether various moments matched.  In particular, we compared the empirical mean, standard deviation, and proportion of zeros to the corresponding statistics which would be predicted by the learned models.

\includegraphics[width=0.8\textwidth]{pics/zinbfitmatch}

In terms of mean and zero proportion, these seem nearly okay, but in terms of variance we can see that we are not doing the greatest job.  For a great many cell types we are predicting standard deviations which are much higher than the reality.  This is perhaps not surprising, since the negative binomial model is inherently overdispersed, and simply cannot accomodate underdispersed data.

Another possible source of trouble is that in the particular model we use we have made certain assumption about how $\alpha,\beta,h$ can vary.  This could be a big problem.  To see why, we looked at each the fitted parameters of the zero-inflated negative binomial distribution for each cell-type.  This yields a kind of three-dimensional scatterplot, in which each dot is a cell type, and the location indicates the value of each of its three fitted parameters.  Here we show three projections of that scatterplot:

\includegraphics[width=0.8\textwidth]{pics/zinbfitdiversity}

Clearly there is quite a lot of variety in terms of how these parameters may vary.  They do not appear to lie on any kind of low-dimensional manifold.  So we must be very careful in terms of what assumptions our model makes about how these parameters may be tied together.

### 2018-03-29 trygve/outline.tex
new residual plots

There is one possibly confusing aspect of our output that we would like to to address.  The upper and lower bounds (as we have visualized them) are each actually impossible as values for $q$.  For example, the rows of the upper bounds generally sum to more than 1.  This is because we have taken \emph{elementwise bounds} on each entry.  An alternative is to  sample uniformly from the set of possible values of $q$ which are consistent with our observations.  We hope this may give the user another way to understand what possible values $q$ might take on.

To this end, we use a Dikin ellipsoid sampler to approximate uniform samples from the set of valid $q$s which are consistent with what we've observed.  These are shown in Figure \ref{fig:allenavg}.  Whether the MCMC method is mixing or not is of course up for debate.  See Appendix \ref{sec:dikin} for some details.

>>>>

Note that the upper and lower bounds (as we have visualized them) are each actually impossible as values for $q$.  For example, the rows of the upper bounds generally sum to more than 1.  This is because we have taken \emph{elementwise bounds} on each entry.  An alternative is to  sample uniformly from the set of possible values of $q$ which are consistent with our observations.  We hope this may give the user another way to understand what possible values $q$ might take on.  To this end, we use a Dikin ellipsoid sampler to approximate uniform samples from the set of valid $q$s which are consistent with what we've observed.  These are shown in Figure \ref{fig:allenavg}.  Whether the MCMC method is mixing or not is of course up for debate.  See Appendix \ref{sec:dikin} for some details.

Another way to look at the space of possible values of $q$ which are consistent with the observables is to try to understand the scope of the polytope along various directions.  In this effort, we selected a direction uniformly at random and found the two extremal vertices in this direction.  We then computed the distance between those vertices, as well as the distance between those vertices projected to the direction.  Doing this procedure many times and making a scatterplot yields Figure \ref{fig:allenwidths}.  This figure suggests that the polytope may be only about 8 units in diameter, although we caution that general results on polytope diameter estimation are not encouraging, cf. \citep{brieden1998approximation}.  However, considering that the magnitude of this diameter would be distributed over 1624 entries of the matrix, each of which lies in $[0,1]$, it would seem that the polytope may indeed be constrained to be very close to the output we showed in Figure \ref{fig:allenoutput}.
